On the Social Compatibility of Genome Analysis

### 2022-06-16 draft.tex
sort surces
sort sources
mini review 3

\begin{abstract}
This paper summarises the chances and risks of genome analysis broken down in three fundamental questions and one discussion. \textit{Who are you} focuses on the individual consumer of DNA tests, what is expected to gain from the results and why it is growing in popularity. \textit{What have you done} takes a look at the past of an individual and what information can be tracked by DNA. \textit{What does your future look like} aims at the consequences this information can have with a focus on the medical sector.
\textit{Consequences for society} discusses what might stem from use and misuse of this technology, specifically eugenics with possible consequence of steering the wrong way, dilemmata with parallels to the abortion debate, amplified marginalization of biological minorities and widening the social division, as well as examining genetic discrimination on the example of insurance.
\end{abstract}
\section{Introduction}
A few centuries ago, life love and a the universe were a mystery, but driven by deeply rooted human curiosity we are beginning to untangle the mess and understand the foundation that grew consciousness out of thin air. One piece of this puzzle is the molecule of life, DeoxyriboNucleic Acid. 
The genetic material that after being refined by millennia of pure chance and survival holds the blueprint to every living creature on our planet. With growing understanding about its structure and the technological advancements that gave us tools to manipulate it, humanity has set its foot in the doorstep to a world in which DNA is no different than a LEGO brick and nothing is out of reach.
This paper aims to explore aspects of how applied genetics will impact us in regard to ethics and data privacy. 
Since great responsibility comes with great power, Genetic engineering is as much blessing as curse and must be used with caution, while genome analysis unveils a tremendous amount of information not just about us, but about our relatives as well and enables more and more accurate predictions about the future.

@@

To recapitulate this approximate timeline should give an overview

\vspace{0.5cm}
	\startchronology[startyear=-2000, stopdate=1850,arrowheight=0.2cm,arrowwidth=0.2cm,width=8cm,height=0.05cm]
	\chronoevent{1856}{Mendel}
	\chronoevent{-370}{Hippocrates \& Aristotle}
	\chronoevent{-2000}{Assyrians}
	\stopchronology

	\startchronology[startyear=1920, stopdate=2022,arrowheight=0.2cm,arrowwidth=0.2cm,width=8cm,height=0.05cm]
	\chronoevent{2022}{This Paper}
	\chronoevent[markdepth=120pt]{2011}{Successful CRISPR Application}
	\chronoevent[markdepth=80pt]{2003}{HGP Completion}
	\chronoevent[markdepth=40pt]{1996}{Cloning of Dolly}
	\chronoevent{1987}{CRISPR Discovery}
	\chronoevent[markdepth=40pt]{1944}{A-Ml-Mc Experiment}
	\chronoevent{1928}{Griffiths Experiment}

	\stopchronology
\vspace{0.5cm}
It's important to keep in mind that current methods operate on single cells, which while passing the modified DNA on during mitosis, are clearly outnumbered by non-modified cells\cite{crispr}.
This can be worked around in three ways.
\begin{itemize}
\item Modify a sufficient number of cells (most likely infeasible for broad changes)
\item Modify in an early stage of development, where only a number of cells exist
\item Germline editing, which means editing the parents germ cells (cells involved in reproduction) in way that the offspring inherits the changes.
\end{itemize}
The third approach differs from the other (so called somatic / nonheritable) techniques in that it can irreversibly affect future generations of offspring with undoubtedly vast consequences, which is why there is widespread advocation to refrain from practicing it in humans\cite{dontedit}.

@@

\section{Who Are You}

\subsection{DNA and Identity}
\textit{Who am I} is a question of humankind. Many try to find an answer to it - and if not, they will be confronted with opinions of others without being asked. Sociologist Abels describes identity as the answer to this question. 
In addition to that, he proposes that in advanced modernity we are increasingly asked the question of our own identity. 
As a result, people start to wonder whether they really know or want to know everything about themselves, or whether they have the courage and strength to face uncomfortable truths. 
According to Abels, identity is what distinguishes individuals from one another\cite{abelsidentity}.

Identity is a concept in numerous disciplines, from psychology and pedagogy to sociology, ethnology, social and cultural anthropology, historical and literary studies to philosophy, and beyond that in many trans- and interdisciplinary debates, and has been for decades. 
To pursue all approaches would require more than a separate paper.

@@

This chapter focuses on online providers of direct-to-customer DNA testing for the following two reasons: First, marketing activities are easy to view and analyze on the websites. Secound, the terms of use of online providers are listed on the website. The investigation is limited to non-health-related DNA testing as further chapters go into more detail on the healthcare sector.

@@

\subsection{Forensics}
Fingerprints, blood splatter, hair, skin flake, salvia, and other fluids – the human body leaves traces everywhere. Therefore, its materiality has been a key part of criminal investigations throughout history. 
With the progression of forensic genetics in the last 35 years, many crime offenders have been identified because of the ability to extract DNA profiles\cite{machado}.

@@

\section{Consequences for Society: Social Compatibility}

@@

\subsection{Analysis}
Analogous to any piece of sensitive and non-public information, knowledge derived from genetic analysis is prone to being exploited as soon as conflicting interest arises.
Companies refusing to employ people who lost part of the genetic lottery would undoubtedly be unfair.
Even more so with technology being fallible. Why should someone whose condition is yet to be diagnosed be more fit for a job than someone with a genetic condition that can already be tested for?

@@

@@

\vspace{5cm}
\begin{thebibliography}{99}

>>>>

\begin{abstract}
This paper summarises the chances and risks of genome analysis broken down in three fundamental questions and one discussion. \textit{Who are you} focuses on the individual consumer of DNA tests, what is expected to gain from the results and why it is growing in popularity. \textit{What have you done} takes a look at the past of an individual and what information can be tracked by DNA. \textit{What does your future look like} aims at the consequences this information can have with a focus on the medical sector.
\textit{Consequences for society} discusses what might stem from use and misuse of this technology, specifically eugenics with possible consequence of steering the wrong way, dilemmata with parallels to the abortion debate, amplified marginalization of biological minorities and widening the social division, as well as examining genetic discrimination on the example of insurance and covering the commercial utilization of this new methodology.
\end{abstract}
\section{Introduction}
A few centuries ago, life love and a the universe were a mystery, but driven by deeply rooted human curiosity we are beginning to untangle the mess and understand the foundation that grew consciousness out of thin air. One piece of this puzzle is the molecule of life, DeoxyriboNucleic Acid. 
The genetic material that after being refined by millennia of pure chance and survival holds the blueprint to every living creature on our planet. With growing understanding about its structure and the technological advancements that gave us tools to manipulate it, humanity has set its foot in the doorstep to a world in which DNA is no different than a LEGO brick and nothing is out of reach.
This paper aims to explore aspects of how applied genetics will impact us in regard to ethics and data privacy. 
Since great responsibility comes with great power, Genetic engineering is as much blessing as curse and must be used with caution, while genome analysis unveils a tremendous amount of information not just about us, but about our relatives as well and enables more and more accurate predictions about the future.

@@

To recapitulate, this approximate timeline in Figure 1 should give an overview
\vspace{0.5cm}
	\startchronology[startyear=1920, stopdate=2022,arrowheight=0.2cm,arrowwidth=0.2cm,width=8cm,height=0.05cm]
	\chronoevent{2022}{This Paper}
	\chronoevent[markdepth=120pt]{2011}{Successful CRISPR Application}
	\chronoevent[markdepth=80pt]{2003}{HGP Completion}
	\chronoevent[markdepth=40pt]{1996}{Cloning of Dolly}
	\chronoevent{1987}{CRISPR Discovery}
	\chronoevent[markdepth=40pt]{1944}{A-Ml-Mc Experiment}
	\chronoevent{1928}{Griffiths Experiment}
	\stopchronology
It's important to keep in mind that current methods operate on single cells, which while passing the modified DNA on during mitosis, are clearly outnumbered by non-modified cells\cite{crispr}.
This can be worked around in three ways.
\begin{itemize}
\item Modify a sufficient number of cells (most likely infeasible for broad changes)
\item Modify in an early stage of development, where only a number of cells exist
\item Germline editing, which means editing the parents germ cells (cells involved in reproduction) in way that the offspring inherits the changes.
\end{itemize}
The third approach differs from the other (so called somatic / nonheritable) techniques in that it can irreversibly affect future generations of offspring with undoubtedly vast consequences, which is why there is widespread advocation to refrain from practicing it in humans\cite{dontedit}.

@@

\section{Who Are You}
This chapter covers the connection between DNA and our sense of self, our need for privacy and right to keep genetic information to ourselves.
\subsection{DNA and Identity}
\textit{Who am I} is a question of humankind. Many try to find an answer to it - and if not, they will be confronted with opinions of others without being asked. Sociologist Abels describes identity as the answer to this question. 
In addition to that, he proposes that in advanced modernity we are increasingly asked the question of our own identity. 
As a result, people start to wonder whether they really know or want to know everything about themselves, or whether they have the courage and strength to face uncomfortable truths. 
According to Abels, identity is what distinguishes individuals from one another\cite{abelsidentity}.

Identity is a concept in numerous disciplines, from psychology and pedagogy to sociology, ethnology, social and cultural anthropology, historical and literary studies to philosophy, and beyond that in many trans- and interdisciplinary debates, and has been for decades. 
To pursue every perspective would require more than a separate paper.

@@

@@

Fingerprints, blood splatter, hair, skin flake, salvia, and other fluids – the human body leaves traces everywhere. Therefore, its materiality has been a key part of criminal investigations throughout history. 
With the progression of forensic genetics in the last 35 years, many crime offenders have been identified because of the ability to extract DNA profiles\cite{machado}.

@@

\section{Consequences for Society: Social Compatibility}
Genetic testing has already established itself in the developed world. This chapter deals with the expected consequences of widespread application of analysis and one day manipulation tools, as well as the commercial exploitation of the underlying technology.
\subsection{Analysis}
Analogous to any piece of sensitive and non-public information, knowledge derived from genetic analysis is prone to being exploited as soon as conflicting interest arises.
Companies refusing to employ people who lost part of the genetic lottery would undoubtedly be unfair.
Even more so with technology being fallible. Why should someone whose condition is yet to be diagnosed be more fit for a job than someone with a genetic condition that can already be tested for?

@@

@@

This section focuses on online providers of direct-to-customer DNA testing for the following two reasons: First, marketing activities are easy to view and analyze on the websites. Secound, the terms of use of online providers are listed on the website. The investigation is limited to non-health-related DNA testing as further chapters go into more detail on the healthcare sector.

@@

\begin{thebibliography}{99}
